---
output: github_document
bibliography: biblio.bib 
csl: 2d-materials.csl
---


```{r, results = 'hide', warning = FALSE, error = FALSE, message = FALSE, echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# hdrm

<!-- badges: start -->
<!-- badges: end -->

R package for performing tests on high dimensional repeated measure data for one group @Pauly2015-jo or multiple groups @Sattler2018-az.

## Installation

The current version can be installed with:

``` {r installation, results = 'hide', warning = FALSE, error = FALSE, message = FALSE}
## install devtools package
if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools")
}
# install package
devtools::install_github("Schnieboli/hdrm", dependencies = TRUE)
```

## One group case

A one group test can be performed by using the function hdrm1. Generic functions exist for the classes matrix, list and data.frame.

```{r example hdrm1, results = 'hide', warning = FALSE, error = FALSE, message = FALSE}
library(hdrm)
## hdrm1 with matrix
N <- 60
d <- 30
# example data with N = 60 subjects in columns and d = 30 factor levels in rows
M <- matrix(rnorm(d*N), d, N)

# test M for H_0 = mu_1 = mu_2 = mu_3 = 0 (no effect in first three factor levels)
H <- matrix(0,d,d)
H[c(1, d+2, 2*d+3)] <- 1
hdrm1(data = M, hypothesis = H)

## hdrm1 with data.frame in longtable format
head(EEG)
# for 'formula' we are interested in value = value, subject = subject and factor = dimension
formula <- value ~ subject + dimension

# test EEG for equal time profile
hdrm1(data = EEG, formula = formula, hypothesis = "equal")


```


## Multiple group case
A test for multiple groups can be performed by the function hdrm_test. Here also exist generics for objects of class matrix, list and data.frame.
```{r example hdrm_test, results = 'hide', warning = FALSE, error = FALSE, message = FALSE}
library(hdrm)
## hdrm1 with matrix
N <- 60
d <- 30
# example data with N = 60 subjects in columns and d = 30 factor levels in rows...
M <- matrix(rnorm(d*N), d, N)

# ... divided into a = 6 groups of n_i = 10
group <- rep(1:6, each = 10)

# test M for group differences
#hdrm_test(data = M, hypothesis = "time", group = group)

## hdrm1 with data.frame in longtable format
head(EEG)
# for 'formula' we are interested in value = value, subject = subject and factor = dimension
formula <- value ~ subject + dimension

# test EEG for equal time profile
hdrm1(data = EEG, formula = formula, hypothesis = "equal")


