---
output: github_document
bibliography: inst/REFERENCES.bib 
csl: 2d-materials.csl
---


```{r, results = 'hide', warning = FALSE, error = FALSE, message = FALSE, echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# hdrm

<!-- badges: start -->
[![DOI](https://img.shields.io/badge/DOI-10.48550%2FarXiv.2512.17478-blue)](https://doi.org/10.48550/arXiv.2512.17478)
[![R-CMD-check](https://github.com/Schnieboli/hdrm-2026/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Schnieboli/hdrm-2026/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The `hdrm` package provides inference procedures for high-dimensional
repeated-measures data in one-sample and multiple-group designs
[@Pauly2015; @Sattler2018].

## Installation

The current version can be installed with:

```{r installation, eval = FALSE, results = 'hide', warning = FALSE, error = FALSE, message = FALSE}
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}

remotes::install_github(
  "Schnieboli/hdrm",
  dependencies = TRUE
)
```

Because `hdrm` contains compiled C++ code, Windows users installing from source need the `Rtools` version corresponding to their version of `R`.

## Citation

When using `hdrm` in a scientific publication, please cite this article @SattlerHichert2025hdrm. The citation can also be obtained directly in `R`:

```{r citation, eval = FALSE}
citation("hdrm")
```

## Data formats

Both main functions accept either:

- a numeric matrix with subjects in rows and repeated-measurement dimensions in columns, or
- a data.frame with columns `value`, `subject` and `dimension`, giving the measurements and the subject and dimension IDs.

In the grouped case, a vector specifying which subject belongs to which group must be passed to `group`.

Missing values in `data` (and `group`) are not allowed and will result in an error. At least two dimensions are required. The one-group method requires at least three subjects, while in the multi-group case, each group must contain at least six complete subjects.

## One-Group Test

A one-group test can be performed using `hdrm_single()`. The functionaccepts either a numeric matrix with dimensions in rows and subjects incolumns, or a data.frame containing columns `value`, `subject` and `dimension`.

The package includes two example data sets: `birthrates` @birthrates,
a wide-format data set containing birth rates for the 16 German federal
states from 1990 to 2023, and `EEG` @EEG_dataset, a long-format data set
containing EEG measurements of 160 individuals in 40 dimensions.

### Wide matrix input

```{r example-hdrm-single-matrix, eval = FALSE, results = 'hide', warning = FALSE, error = FALSE, message = FALSE}
data("birthrates")

## transform 'birthrates' to matrix and transpose
birthrates_matrix <- t(as.matrix(birthrates))

## test whether the time profile is flat
hdrm_single(
  data = birthrates_matrix,
  hypothesis = "flat"
)

## define hypothesis = "flat" via equivalent projection matrix
d <- ncol(birthrates_matrix)
flat_projection <- diag(d) -
  matrix(
    1 / d,
    nrow = d,
    ncol = d
  )

## test whether the time profile is flat via custom hypothesis matrix
hdrm_single(
  data = birthrates_matrix,
  hypothesis = flat_projection
)
```

### Long data.frame input

The included `EEG` data contain several diagnostic groups. The following
example selects one group for a one-group analysis.

```{r example-hdrm-single-data.frame, eval = FALSE, results = 'hide', warning = FALSE, error = FALSE, message = FALSE}
data("EEG")
names(EEG) # already contains columns value, dimension and subject

## select only one diagnostic group for one-group analysis
EEG_single <- EEG[EEG$group == "SCC+", ]

## test whether the time profile is flat
hdrm_single(
  data = EEG_single,
  hypothesis = "flat"
)

## define hypothesis = "flat" via equivalent projection matrix
d <- nlevels(EEG_single$dimension)
flat_projection <- diag(d) -
  matrix(1 / d, nrow = d, ncol = d)

## test whether the time profile is flat via custom hypothesis matrix
hdrm_single(
  data = EEG_single,
  hypothesis = flat_projection
)
```

## Multiple-Group inference

`hdrm_grouped()` covers two different covariance settings.

| Setting | Main arguments | Interpretation of `B` |
|---|---|---|
| Heterogeneous covariances with exact available trace estimators | `cov.equal = FALSE`, `subsampling = FALSE` | Base budget; the joint third-trace estimator uses `a * B` joint draws |
| Heterogeneous covariances with additional subsampling | `cov.equal = FALSE`, `subsampling = TRUE` | Base budget; group-specific and pairwise estimators use `B` draws per group or pair, while the joint third-trace estimator uses `a * B` joint draws |
| Equal covariance matrices | `cov.equal = TRUE` | Base budget; the pooled third-trace estimator uses `a * B` total draws distributed across groups |

For the equal-covariance method, `subsampling` has no effect. The `a * B`
third-trace draws are allocated across groups approximately proportionally to
the numbers of available six-subject subsets.

A third-trace quantity is estimated by subsampling in every grouped analysis.
Consequently, `f`, `tau`, and the p-value depend on `B` and the random seed even
when `subsampling = FALSE`. With heterogeneous covariances and
`subsampling = TRUE`, the test statistic itself is seed dependent as well.

Supplying `seed` makes a call reproducible without permanently changing the
previous `R` random-number state.


```{r example-hdrm-grouped, eval = FALSE, results = 'hide', warning = FALSE, error = FALSE, message = FALSE}
library(hdrm)

### Multiple-group test for data in wide-table format using the built-in birthrates data set

data("birthrates", package = "hdrm")

# A group vector with one entry for each subject, here each federal state,
# is required. The states are divided into western and eastern states.
group <- factor(
  c(1, 1, 2, 2, 1, 1, 1, 2, 1, 1, 1, 1, 2, 2, 1, 2),
  labels = c("west", "east")
)

# For matrix data, dimensions are stored in rows and subjects in columns.
hdrm_grouped(
  data = as.matrix(birthrates),
  hypothesis = "interaction",
  group = group
)

### Multiple-group test for data in long-table format using the built-in EEG data set

data("EEG", package = "hdrm")

# Construct subject identifiers that are unique across groups.
eeg_grouped <- EEG
eeg_grouped$subject_id <- interaction(
  eeg_grouped$group,
  eeg_grouped$subject,
  drop = TRUE
)

# Sort the observations so that the dimensions occur in a consistent order
# for every subject.
eeg_grouped <- eeg_grouped[
  order(
    eeg_grouped$group,
    eeg_grouped$subject_id,
    eeg_grouped$dimension
  ),
]

# For long-format data, pass the measurements, group identifiers,
# and subject identifiers as separate vectors.
hdrm_grouped(
  data = eeg_grouped$value,
  hypothesis = "whole",
  group = eeg_grouped$group,
  subject = eeg_grouped$subject_id
)
```

## References
