---
output: github_document
bibliography: inst/REFERENCES.bib 
csl: 2d-materials.csl
---


```{r, results = 'hide', warning = FALSE, error = FALSE, message = FALSE, echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# hdrm

<!-- badges: start -->
[![DOI](https://img.shields.io/badge/DOI-10.48550%2FarXiv.2512.17478-blue)](https://doi.org/10.48550/arXiv.2512.17478)
<!-- badges: end -->

The `hdrm` package provides inference procedures for high-dimensional
repeated-measures data in one-sample and multiple-group designs
[@Pauly2015; @Sattler2018].

## Installation

The current version can be installed with:

```{r installation, eval = FALSE, results = 'hide', warning = FALSE, error = FALSE, message = FALSE}
## install devtools package
if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools")
}
# install package
devtools::install_github("Schnieboli/hdrm", dependencies = TRUE)
```
Be aware that the respective [rtools-version](https://cran.r-project.org/bin/windows/Rtools/) for your version of R is required for installation when using Windows.

## Citation

If you use `hdrm` in a scientific publication, please cite
@SattlerHichert2025hdrm.

The citation can also be obtained in R using:

```{r citation, eval = FALSE}
citation("hdrm")
```

## One-Group Test

A one-group test can be performed using `hdrm_single()`. The function
accepts either a numeric matrix with dimensions in rows and subjects in
columns, or a numeric vector together with a subject identifier.

The package includes two example data sets: `birthrates` @birthrates,
a wide-format data set containing birth rates for the 16 German federal
states from 1990 to 2023, and `EEG` @EEG_dataset, a long-format data set
containing EEG measurements in 40 dimensions.

```{r example-hdrm-single, eval = FALSE, results = 'hide', warning = FALSE, error = FALSE, message = FALSE}
library(hdrm)

### One-group test for data in wide format

data("birthrates", package = "hdrm")

hdrm_single(
  data = as.matrix(birthrates),
  hypothesis = "flat"
)

### One-group test for data in long-table format using the built-in EEG data set

data("EEG", package = "hdrm")

# The EEG data contain observations from multiple groups.
# For a one-group test, select the first group.
eeg_single <- droplevels(
  EEG[EEG$group == levels(EEG$group)[1], ]
)

# Sort the observations by subject and dimension so that the measurements
# occur in the same dimensional order for every subject.
eeg_single <- eeg_single[
  order(eeg_single$subject, eeg_single$dimension),
]

# A hypothesis can also be specified through a matrix.
# This matrix is equivalent to the predefined hypothesis "flat".
d <- nlevels(eeg_single$dimension)
hypothesis <- diag(d) - matrix(1 / d, nrow = d, ncol = d)

# For long-format data, pass the numeric measurements as a vector and
# provide the corresponding subject identifiers separately.
hdrm_single(
  data = eeg_single$value,
  hypothesis = hypothesis,
  subject = eeg_single$subject
)
```

## Multiple-Group Test

A multiple-group test can be performed using `hdrm_grouped()`. The
function accepts either a numeric matrix with dimensions in rows and
subjects in columns, or a numeric vector together with subject and group
identifiers.

The test uses subsampling-based estimators for computationally demanding
quantities. The number of subsamples can be controlled through the
argument `B`. Setting `subsampling = TRUE` additionally uses the
subsampling versions for the other estimators.

```{r example-hdrm-grouped, eval = FALSE, results = 'hide', warning = FALSE, error = FALSE, message = FALSE}
library(hdrm)

### Multiple-group test for data in wide-table format using the built-in birthrates data set

data("birthrates", package = "hdrm")

# A group vector with one entry for each subject, here each federal state,
# is required. The states are divided into western and eastern states.
group <- factor(
  c(1, 1, 2, 2, 1, 1, 1, 2, 1, 1, 1, 1, 2, 2, 1, 2),
  labels = c("west", "east")
)

# For matrix data, dimensions are stored in rows and subjects in columns.
hdrm_grouped(
  data = as.matrix(birthrates),
  hypothesis = "interaction",
  group = group
)

### Multiple-group test for data in long-table format using the built-in EEG data set

data("EEG", package = "hdrm")

# Construct subject identifiers that are unique across groups.
eeg_grouped <- EEG
eeg_grouped$subject_id <- interaction(
  eeg_grouped$group,
  eeg_grouped$subject,
  drop = TRUE
)

# Sort the observations so that the dimensions occur in a consistent order
# for every subject.
eeg_grouped <- eeg_grouped[
  order(
    eeg_grouped$group,
    eeg_grouped$subject_id,
    eeg_grouped$dimension
  ),
]

# For long-format data, pass the measurements, group identifiers,
# and subject identifiers as separate vectors.
hdrm_grouped(
  data = eeg_grouped$value,
  hypothesis = "whole",
  group = eeg_grouped$group,
  subject = eeg_grouped$subject_id
)
```

## References
