---
output: github_document
bibliography: inst/REFERENCES.bib 
csl: apa.csl
---


```{r, results = 'hide', warning = FALSE, error = FALSE, message = FALSE, echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(hdrm)
```
# hdrm

<!-- badges: start -->
<!-- badges: end -->

R package for performing tests on high dimensional repeated measure data for one group [@Pauly2015] or multiple groups [@Sattler2018; @Sattler2025].

## Installation

The current version can be installed by:

``` {r installation, results = 'hide', warning = FALSE, error = FALSE, message = FALSE}
## install devtools package
if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools")
}
# install package
devtools::install_github("Schnieboli/hdrm", dependencies = TRUE)
```
Be aware that the respective [rtools-version](https://cran.r-project.org/bin/windows/Rtools/) for your version of R is required for installation when using Windows.

## Data Sets
The package has two data sets included: `birthrates` [@birthrates] is a matrix that contains the birthrates of German states from 1990 to 2023. `EEG` [@EEG_dataset] is a data frame that contains EEG data in 40 dimensions.

```{r data sets, results = 'hide', warning = FALSE, error = FALSE, message = FALSE}
data("birthrates")
?birthrates

data("EEG")
?EEG
```

## One Group Test

A one group test can be performed by using the function `hdrm_single`. The argument `data` that specifies the data can be a matrix or a vector. If `data`  is  a matrix, it is expected that subjects are represented by columns and factor levels are represented by rows. If `data` is a vector, `subject` must be a vector of the same length that allocates the measurements in `data` to the subjects.

```{r example hdrm_single, results = 'hide', warning = FALSE, error = FALSE, message = FALSE}
library(hdrm)

##### One sample test with built in data set birthrates
hdrm_single(data = birthrates,
            hypothesis = "flat", # test whether time profile is flat
)

##### One sample test with built in data set EEG
# hypothesis can also be given as a matrix
hypothesis <- matrix(1/40, nrow = 40, ncol = 40) 
# matrix equivalent to hypothesis = 'flat'

hdrm_single(data = EEG$value,
            hypothesis = hypothesis,
            subject = EEG$subject,
)
```


## Multiple Group Test
A test for multiple groups can be performed by using the function `hdrm_grouped`. The argument `data` that specifies the data can be a matrix or a vector. If `data` is  a matrix, it is expected that subjects are represented by columns and factor levels are represented by rows. The argument `group` then allocates the subjects to a group. If `data` is a vector, The arguments `subject` and `group` must be vectors of the same length that allocates the measurements in `data` to the subjects and the subjects to a group.

The test is performed using subsampling to estimate the computationally heaviest estimator. The number of subsamples can be controlled via the argument `B`. If your data is large, setting `subsampling = TRUE` results in the subsampling versions for the other estimators being used as well.

```{r example hdrm_grouped, results = 'hide', warning = FALSE, error = FALSE, message = FALSE}
library(hdrm)

##### Test for multiple groups with built in data set birthrates

## divide German states in east and west:
group <- factor(c(1,1,2,2,1,1,1,2,1,1,1,1,2,2,1,2), labels = c("west", "east"))
hdrm_grouped(data = birthrates,
             ## test for interaction effect between group and dimension
             hypothesis = "interaction",
             group = group
)

##### Test for multiple groups with built in data set EEG
# hypothesis can also be given as a matrix
hypothesis <- list(TW = diag(4) - matrix(1/4, 4, 4),
                   TS = matrix(1/40, 40, 40)
) # list entries equivalent to hypothesis = 'whole'

hdrm_grouped(data = EEG$value,
             hypothesis = hypothesis, # test for time effect in groups
             group = EEG$group,
             subject = EEG$subject,
)
```

The parameter `cov.equal` can be set to `TRUE` to perform the test for equal covariances between groups if this assumption can be justified [@Sattler2021]. To reduce computational complexity, an alternative hypothesis matrix can be employed. [@Sattler2025] establish the existence of a companion matrix with minimal row dimension that yields identical values for the considered test statistics.The necessary computational tools to determine these matrices are implemented in the \texttt{HypoShrink R} package [@HypoShrink].

### References
