---
output: github_document
bibliography: inst/REFERENCES.bib 
csl: 2d-materials.csl
---


```{r, results = 'hide', warning = FALSE, error = FALSE, message = FALSE, echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# hdrm

<!-- badges: start -->
[![DOI](https://img.shields.io/badge/DOI-10.48550%2FarXiv.2512.17478-blue)](https://doi.org/10.48550/arXiv.2512.17478)
[![R-CMD-check](https://github.com/Schnieboli/hdrm-2026/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Schnieboli/hdrm-2026/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The `hdrm` package provides inference procedures for high-dimensional
repeated-measures data in one-sample and multiple-group designs
[@Pauly2015; @Sattler2018].

## Installation

The current version can be installed with:

```{r installation, eval = FALSE, results = 'hide', warning = FALSE, error = FALSE, message = FALSE}
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}

remotes::install_github(
  "Schnieboli/hdrm",
  dependencies = TRUE
)
```

Because `hdrm` contains compiled C++ code, Windows users installing from source need the `Rtools` version corresponding to their version of `R`.

## Citation

When using `hdrm` in a scientific publication, please cite this article @SattlerHichert2025hdrm. The citation can also be obtained directly in `R`:

```{r citation, eval = FALSE}
citation("hdrm")
```

## Data formats

Both main functions accept either:

- a numeric matrix with subjects in rows and repeated-measurement dimensions in columns, or
- a data.frame with columns `value`, `subject` and `dimension`, giving the measurements and the subject and dimension IDs.

In the grouped case, a vector specifying which subject belongs to which group must be passed to `group`.

Missing values in `data` (and `group`) are not allowed and will result in an error. At least two dimensions are required. The one-group method requires at least three subjects, while in the multi-group case, each group must contain at least six complete subjects.

## One-Group Test

A one-group test can be performed using `hdrm_single()`. The functionaccepts either a numeric matrix with dimensions in rows and subjects incolumns, or a data.frame containing columns `value`, `subject` and `dimension`.

The package includes two example data sets: `birthrates` @birthrates,
a wide-format data set containing birth rates for the 16 German federal
states from 1990 to 2023, and `EEG` @EEG_dataset, a long-format data set
containing EEG measurements of 160 individuals in 40 dimensions.

### Wide matrix input

```{r example-hdrm-single-matrix, eval = FALSE, results = 'hide', warning = FALSE, error = FALSE, message = FALSE}
data("birthrates")

## transform 'birthrates' to matrix and transpose
birthrates_matrix <- t(as.matrix(birthrates))

## test whether the time profile is flat
hdrm_single(
  data = birthrates_matrix,
  hypothesis = "flat"
)

## define hypothesis = "flat" via equivalent projection matrix
d <- ncol(birthrates_matrix)
flat_projection <- diag(d) -
  matrix(
    1 / d,
    nrow = d,
    ncol = d
  )

## test whether the time profile is flat via custom hypothesis matrix
hdrm_single(
  data = birthrates_matrix,
  hypothesis = flat_projection
)
```

### Long data.frame input

The included `EEG` data contain several diagnostic groups. The following
example selects one group for a one-group analysis.

```{r example-hdrm-single-data.frame, eval = FALSE, results = 'hide', warning = FALSE, error = FALSE, message = FALSE}
data("EEG")
names(EEG) # already contains columns value, dimension and subject

## select only one diagnostic group for one-group analysis
EEG_single <- EEG[EEG$group == "SCC+", ]

## test whether the time profile is flat
hdrm_single(
  data = EEG_single,
  hypothesis = "flat"
)

## define hypothesis = "flat" via equivalent projection matrix
d <- nlevels(EEG_single$dimension)
flat_projection <- diag(d) -
  matrix(1 / d, nrow = d, ncol = d)

## test whether the time profile is flat via custom hypothesis matrix
hdrm_single(
  data = EEG_single,
  hypothesis = flat_projection
)
```

## Multiple-Group inference

`hdrm_grouped()` covers two different covariance settings.

| Setting | Main arguments | Interpretation of `B` |
|---|---|---|
| Heterogeneous covariances with exact available trace estimators | `cov.equal = FALSE`, `subsampling = FALSE` | Base budget; the joint third-trace estimator uses `a * B` joint draws |
| Heterogeneous covariances with additional subsampling | `cov.equal = FALSE`, `subsampling = TRUE` | Base budget; group-specific and pairwise estimators use `B` draws per group or pair, while the joint third-trace estimator uses `a * B` joint draws |
| Equal covariance matrices | `cov.equal = TRUE` | Base budget; the pooled third-trace estimator uses `a * B` total draws distributed across groups |

For the equal-covariance method, `subsampling` has no effect. The `a * B`
third-trace draws are allocated across groups approximately proportionally to
the numbers of available six-subject subsets.

A third-trace quantity is estimated by subsampling in every grouped analysis.
Consequently, `f`, `tau`, and the p-value depend on `B` and the random seed even
when `subsampling = FALSE`. With heterogeneous covariances and
`subsampling = TRUE`, the test statistic itself is seed dependent as well.

Supplying `seed` makes a call reproducible without permanently changing the
previous `R` random-number state.


### Wide matrix input

```{r example-hdrm-grouped-matrix-1, eval = FALSE, results = 'hide', warning = FALSE, error = FALSE, message = FALSE}
data("birthrates")

birthrates_matrix <- t(as.matrix(birthrates))

# group states into east and west (Berlin as east)
group <- factor(c(1, 1, 2, 2, 1, 1, 1, 2, 1, 1, 1, 1, 2, 2, 1, 2), 
                labels = c("west", "east")) 

## test for interaction effect of group and time
hdrm_grouped(
  data = birthrates_matrix,
  hypothesis = "interaction",
  group = group,
  cov.equal = FALSE,
  subsampling = FALSE,
  B = "100*N",
  seed = 3141
)

```

The equal-covariance method is selected explicitly:

```{r example-hdrm-grouped-matrix-2, eval = FALSE, results = 'hide', warning = FALSE, error = FALSE, message = FALSE}
## test for interaction effect of group and time with equal covariance assumption
hdrm_grouped(
  data = birthrates_matrix,
  hypothesis = "interaction",
  group = group,
  cov.equal = TRUE,
  B = "100*N",
  seed = 3141
)

```

### Long data.frame input

If `data` is a data.frame, it must contain columns `value`, `subject` and `dimension`.

```{r example-hdrm-grouped-data.frame, eval = FALSE, results = 'hide', warning = FALSE, error = FALSE, message = FALSE}
data("EEG")
names(EEG) # already contains columns 'value', 'subject' and 'dimension'

## test for differences between the diagnostic groups
hdrm_grouped(
  data = EEG,
  hypothesis = "whole",
  group = EEG$group,
  cov.equal = FALSE,
  subsampling = FALSE,
  B = "100*N",
  seed = 3141
)
```

### Hypotheses

For `hdrm_grouped()`, the predefined hypotheses are:

- `"whole"`: no whole-plot or group main effect,
- `"sub"`: no subplot or dimension main effect,
- `"interaction"`: no group-by-dimension interaction,
- `"identical"`: identical expectation vectors across groups,
- `"flat"`: a flat expectation profile within every group.

A custom grouped hypothesis is supplied as a named list containing projection
matrices `TW` and `TS`:

```{r example-hdrm-grouped-hypothesis.as.list, eval = FALSE, results = 'hide', warning = FALSE, error = FALSE, message = FALSE}
## custom hypothesis for testing for effect of time
custom_hypothesis <- list(
  TW = matrix(1 / 2, nrow = 2, ncol = 2),
  TS = diag(34) - matrix(1 / 34, nrow = 34, ncol = 34)
)

## testing for time effect (equivalent to hypothesis = "sub")
hdrm_grouped(
  data = birthrates_matrix,
  hypothesis = custom_hypothesis,
  group = group,
  B = "100*N",
  seed = 3141
)
```

## Reported p-values

Upper-tail probabilities are calculated directly. To avoid reporting an exact numerical zero, p-values smaller than machine precision are returned as `.Machine$double.eps`. Such a value should be interpreted as being no larger than the numerical reporting threshold.


## References
